DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models

### 2017-04-28 DCMIP2016-Part1-Models.tex
V4 update

\title{DCMIP2016:  A Review of Nonhydrostatic Dynamical Core Design and Intercomparison of Participating Models}

@@

\runningtitle{DCMIP2016: A Review of Nonhydrostatic Dynamical Core Design and Intercomparison}

@@

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind velocity \\
$v$ & Meridional wind velocity \\
$w$ & Vertical wind velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\section{Dynamical Cores} \label{sec:DynamicalCores}

@@

\begin{table}[p]
\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FV$^3$ & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FVM & Co-located & fixed height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
HOMME & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}

@@

\subsection{High-Order Method Modeling Environment (HOMME)}

The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following pressure coordinate in the vertical. The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions. Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.

%A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$. The PDEs are discretized using a split approximation, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

@@

The ICON model \citep{zangl2014icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.  Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.  Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

@@

\section{Model Grids} \label{sec:ModelGrids}

@@

\begin{figure}[]
\includegraphics[height=6cm]{gem_yinyang.png}
\caption{Th Yin-Yang grid, a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
\label{fig:gem_yinyang}
\end{figure}

@@

\subsection{Vertical Coordinates} \label{sec:VerticalCoordinates}

@@

The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
\begin{align} \label{eq:SurfacePressureEquation}
\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
\end{align} where the integrals are taken over the column.

@@

\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and HOMME)} \label{sec:FloatingLagrangian}

@@

\subsection{Prognostic equations} \label{sec:PrognosticEquations}

The Navi\'er-Stokes equations that govern atmospheric motion can take on many forms, depending on the choice of prognostic variables and coordinate system.  A derivation of many forms of these equations can be found in Appendix \ref{sec:EquationSets}.  The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.  The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models.

@@

%GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$, described in section \ref{sec:GEM_zeta}.
In GEM the non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
\begin{align}
\mu = \pdiff{p}{s} - 1,
\end{align}
\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based}).  In this case the equations of GEM model \citep{Girard2014} are concisely given by
\begin{align}
\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
\frac{d w}{dt}- g \mu &= 0, \\
\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
\frac{d \Phi}{dt}-  g w & = 0.
\end{align}

@@

\subsubsection{HOMME}

HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics. The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure (\ref{eq:SurfacePressureEquation}).

In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model. This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}. In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical. Prognostic variables include the 3d wind velocity components $u,v,w$ ((\ref{eq:PrognosticWEquationK2Sb}) and (\ref{eq:PrognosticUhEquationK2Sb})), the surface pressure $p_s$ (\ref{eq:SurfacePressureEquation}), potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).

%Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.

% The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios. Tracer advection is discretized in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}. The dynamics are also discretized using one of several explicit Runge-Kutta times schemes. Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.

%This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability. Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.

@@

\subsection{HOMME / Tempest}

>>>>

\title{DCMIP2016:  A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models}

@@

\runningtitle{DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison}

@@

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind velocity \\
$v$ & Meridional wind velocity \\
$w$ & Vertical wind velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
$\theta_\rho$ & Density potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Specific humidity \\
$q_v$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\section{Dynamical cores} \label{sec:DynamicalCores}

@@

\begin{table}[p]
\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FV$^3$ & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FVM & Co-located & fixed height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
ACME--A & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Global Environmental Multiscale (GEM) model} \label{sec:GEM_core}

@@

\subsection{Accelerated Climate Model for Energy--Atmosphere (ACME--A)}

The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}. ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model. Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed. The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}. All fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions. Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.

%The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME--A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following pressure coordinate in the vertical. The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions. Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.

@@

The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.  Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.  Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

@@

\section{Model grids} \label{sec:ModelGrids}

@@

\begin{figure}[]
\includegraphics[height=6cm]{gem_yinyang.png}
\caption{The Yin-Yang grid is a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
\label{fig:gem_yinyang}
\end{figure}

@@

\subsection{Vertical coordinates} \label{sec:VerticalCoordinates}

@@

%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
%\begin{align} \label{eq:SurfacePressureEquation}
%{\color{red}\pdiff{p_s}{t} = - \int g \pdiff{\rho}{t} \left( \pdiff{z}{s} \right) ds = - \int \nabla_s \cdot (\rho_s \vb{u}_h) \left( \pdiff{z}{s} \right) ds,}
%\end{align} where the integrals are taken over the column.

@@

\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and ACME--A)} \label{sec:FloatingLagrangian}

@@

\subsection{Prognostic equations and treatment of moisture} \label{sec:PrognosticEquations}

The Navi\'er-Stokes equations that govern atmospheric motion can take on many forms, depending on the choice of prognostic variables and coordinate system.  A derivation of many forms of these equations can be found in Appendix \ref{sec:EquationSets}.  The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.  The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models, including any special treatment of terms related to moist physics.

@@

%GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$, described in section \ref{sec:GEM_zeta}.
In GEM the non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
\begin{align}
\mu = \pdiff{p}{s} - 1,
\end{align}
\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based}).  In this case the equations of GEM model \citep{Girard2014} are concisely given by
\begin{align}
\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
\frac{d w}{dt}- g_c \mu &= 0, \label{eq:GEMVerticalVelocity} \\
\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
\frac{d \Phi}{dt}-  g_c w & = 0.
\end{align}

@@

\subsubsection{ACME--A}

ACME-A hosts an experimental compressible shallow-atmosphere model in hybrid terrain-following pressure vertical coordinates $\eta$, similar to the model of \cite{laprise1992euler}. The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), pseudodensity (\ref{eq:PrognosticPseudoDensity}), and geopotential (\ref{eq:PrognosticShallowGeopotential}).  The vertical velocity equation is formulated analogous to that of GEM (\ref{eq:GEMVerticalVelocity}).

@@

\subsection{ACME--A / Tempest}
